Practical Optional Types for Clojure

### 2015-02-19 formal-model.tex
smaller figures
isa figures
type hinting figure
add rewrite figure
type hint inference
tweak type hinting figure
java interop semantics
type hint prose

We describe our type system in a similar way to Typed Racket (\citet{TF10}),
with differences highlighted in $\trdiff{\text{blue}}$.
The main judgement

$
{\judgement   {\propenv}
              {\hastype {\e{}} {\t{}}}
  {\filterset {\thenprop {\prop{}}}
              {\elseprop {\prop{}}}}
  {\object{}}}
$

says expression \e{} is of type \t{} in the 
proposition environment $\propenv{}$, with 
then proposition {\thenprop {\prop{}}}, else proposition {\elseprop {\prop{}}}
and object \object{}.

@@

@@

\subsection{Extending occurrence typing}

@@

@@

Clojure comes with a small type system that helps improve
performance in Clojure code: type hints. They are used to help
the compiler resolve Java methods and prevent unboxing in
certain situations. Tags are propagated bidirectionally
and the programmer is free to add them where they like.

@@

>>>>

We describe our type system in a similar way to Typed Racket (\citet{TF10}),
with differences highlighted in $\trdiff{\text{blue}}$.
The typing judgement 
$
{\judgement   {\propenv}
              {\hastype {\e{}} {\t{}}}
  {\filterset {\thenprop {\prop{}}}
              {\elseprop {\prop{}}}}
  {\object{}}}
$
says expression \e{} is of type \t{} in the 
proposition environment $\propenv{}$, with 
`then' proposition {\thenprop {\prop{}}}, `else' proposition {\elseprop {\prop{}}}
and object \object{}.

The propositions inform conditionals on what they can assume down the then and else
branches if the expression is used as a test.
The object is a possibly-empty symbolic representation of the expression.

\begin{figure*}
  \footnotesize
$$
\begin{altgrammar}
  \expd{}, \e{} &::=& \x{}
                      \alt \v{} 
                      \alt {\comb {\e{}} {\e{}}} 
                      \alt {\abs {\x{}} {\t{}} {\e{}}}
                      \alt {\ifexp {\e{}} {\e{}} {\e{}}}
                      \alt {\trdiff{\doexp {\e{}} {\e{}}}}
                      \alt {\letexp {\x{}} {\e{}} {\e{}}}
                &\mbox{Expressions} \\
  \v{} &::=&          \true{} \alt \false{} \alt \trdiff{\nil{}}
                      \alt \trdiff{\class{}}
                      \alt \k{}
                      \alt {\const{}}
                      \alt {\closure {\openv{}} {\abs {\x{}} {\t{}} {\e{}}}}
                &\mbox{Values} \\
  {\const{}}           &::=& \classconst \alt \throwconst

                &\mbox{Constants}\\
  \s{}, \t{}    &::=& \Top 
                      \alt {\Unionsplice {\overrightarrow{\t{}}}}
                      \alt
                      {\ArrowOne {\x{}} {\t{}}
                                   {\t{}}
                                   {\filterset {\prop{}} {\prop{}}}
                                   {\object{}}}
                      \alt {\Value \singletonmeta{}} 
                &\mbox{Types} \\
  \singletonallsyntax{}
                \\ \\
  \occurrencetypingsyntax{}
  %\\
  %\classliteralallsyntax{}
\end{altgrammar}
$$
\caption{Syntax of Terms, Types, Propositions and Objects}
\end{figure*}

\begin{figure*}
  \footnotesize
\begin{mathpar}
  {\TLocal}

  {\TTrue}

{\TFalse}

{\TNil}

{\TDo}

{\TIf}

{\TLet}

{\TApp}

{\TAbs}

         {\TSubsume}
\end{mathpar}
\caption{Standard typing rules}
\end{figure*}

\begin{figure*}
  \footnotesize
\begin{mathpar}
  {\BLocal}

  {\BDo}

  {\BLet}

  \BVal{}

\BIfTrue{}

\BIfFalse{}

\BAbs{}

\BBetaClosure{}

\BDelta{}
\end{mathpar}
\caption{Semantics}
\end{figure*}

\begin{figure*}
  \footnotesize
\begin{mathpar}
\standardsubtyping{}
\end{mathpar}
\caption{Subtyping rules}
\end{figure*}

\subsection{\isaliteral{}}

One interesting Clojure function we modeled is \isaliteral.
It takes two arguments, and returns \true{} if they are the same (equality mode)
or if they are both classes and the left is a subtype of the right (subclassing mode),
otherwise returns \false{}, as described in figure~\ref{figure:isasemantics}.

For example, {\isaapp{\Keyword}{\Object}} = \true{} but 
{\isaapp{\Object}{\Keyword}} = \false{}.
This is useful in combination with \classconst{}
with a fixed second argument, where {\isaapp{\appexp{\classconst}{\x{}}}{\Keyword}}
is a test for \x{} being a \Keyword{}, and also
as an equality test like
{\isaapp{\x{}}{\makekw{a}}}.

We encode this behaviour in figure~\ref{figure:isatyping}. 
\isacompareliteral{} is the interesting part,
which takes the object of the first argument 
and the type of the second argument, and returns the then and else propositions
of the expression.

The first case handles subclassing mode
in expressions like {\isaapp{\appexp{\classconst}{\x{}}}{\Object}},
where the first argument has an object ending in \classpe{} and the second argument has a singleton class type.
For example, in this case \isacomparetwo{\path{\classpe{}}{\x{}}}{\Value{\Object}}
gives a then proposition {\isprop{\Object} {\x{}}}
and an else proposition {\notprop{\Object}{\x{}}}.

The reason to peel off one path element is subtle. For example,
claiming {\isprop{\Value{\Object}} {\path{\classpe{}}{\x{}}}}
might seem suitable, however it implies \inopenv{\openv{}}{\path{\classpe{}}{\x{}}}{\Object}
which is wrong if \x{} = \makekw{a} where \inopenv{\openv{}}{\path{\classpe{}}{\x{}}}{\Keyword}.
(In formulating \isacompareliteral we found a three year old bug in how core.typed handles this case).

The second case handles the equality mode of \isaliteral{}
with a non-class singleton type as the second argument, to be certain we are not in subclassing mode.
Substitution is used to ensure well formed propositions.

The final case gives up; we cannot distinguish between equality and subclassing mode.

\isaliteral{} serves as the default dispatch mechanism for multimethods in Clojure.

\begin{figure}
  \footnotesize
  \begin{mathpar}
    \isaopsemfigure{}

    \BIsA{}
  \end{mathpar}
  \caption{\isaliteral{} semantics}
  \label{figure:isasemantics}
\end{figure}

\begin{figure}
  \footnotesize
  \begin{mathpar}
    \isapropsfigure{}

    \TIsA{}
  \end{mathpar}
  \caption{\isaliteral{} typing rules}
  \label{figure:isatyping}
\end{figure}

@@

\begin{figure}
  \footnotesize
$$
\begin{altgrammar}
  \e{} &::=& \ldots \alt \multimethodexpsyntax{} 
                &\mbox{Expressions} \\
  \v{} &::=& \ldots \alt {\multi {\v{}} {\disptable{}}}
                &\mbox{Values} \\

 \disptablesyntax{} \\
\end{altgrammar}
$$
\caption{Multimethod Syntax}
\end{figure}

\begin{figure}
  \footnotesize
\begin{mathpar}
  \TDefMulti{}

  \TDefMethod{}
\end{mathpar}
\caption{Multimethod typing rules}
\end{figure}

\begin{figure}
  \footnotesize
\begin{mathpar}
  \Multisubtyping{}
\end{mathpar}
\caption{Multimethod subtyping rules}
\end{figure}

\begin{figure}
  \footnotesize
  \getmethodfigure{}
\begin{mathpar}

       \BDefMulti{}

       \BDefMethod{}

  \BBetaMulti{}
\end{mathpar}
\caption{Multimethod semantics}
\end{figure}

@@

\subsection{Heterogenous maps}

\begin{figure}
  \footnotesize
$$
\begin{altgrammar}
  \e{} &::=& \ldots \alt \hmapexpressionsyntax{}
                &\mbox{Expressions} \\
  \v{} &::=& \ldots \alt {\emptymap{}}
                &\mbox{Values} \\
  \t{} &::=& \ldots \alt {\HMapgeneric {\mandatory{}} {\absent{}}}
                &\mbox{Types} \\
  \auxhmapsyntax{}\\
\end{altgrammar}
$$
\caption{HMap Terms and Types}
\end{figure}

\begin{figure}
  \footnotesize
\begin{mathpar}
  \HMapsubtyping{}
\end{mathpar}
\caption{HMap subtyping rules}
\end{figure}

@@

\begin{figure}
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots \alt \reflectiveexpsyntax{}
    &\mbox{Expressions} \\
    \v{} &::=& \ldots \alt {\classvalue{\classhint{}} {\overrightarrow {\classfieldpair{\fld{}} {\v{}}}}}
    &\mbox{Values} \\
  \end{altgrammar}
  $$
  \caption{Java Interoperability Terms and Types}
\end{figure}

@@

\begin{figure}
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots \alt \mininonreflectiveexpsyntax{}
    &\mbox{Expressions} \\
  \end{altgrammar}
  $$
  \caption{Type Hinting and Non-Reflective Terms}
\end{figure}

\begin{figure*}
  \footnotesize
  \begin{mathpar}
    \RAbs{}

    \RNewElimRefl{}

    \RMethodElimRefl{}

    \RFieldElimRefl{}

    \RLet{}

    \RLetHint{}
  \end{mathpar}
\caption{Reflection Elimination}
\label{figure:rewrite}
\end{figure*}

\begin{figure*}
  \footnotesize
  \begin{mathpar}
    \TALocal{}

    \TANil{}

    \TANewStatic{}

    \TALetHint{}

    \TALet{}
  \end{mathpar}
\caption{Type Hint Inference}
\label{figure:hintinfer}
\end{figure*}

\begin{figure}
  \footnotesize
  \begin{mathpar}
    \BField{}

    \BMethod{}

    \BNew{}
  \end{mathpar}
  \caption{Java Interoperability Semantics}
\end{figure}

Clojure supports \emph{type hints}, an optional type system to improve performance
and disambiguate Java calls.
Programmers communicate to the Clojure compiler via type hints
to resolve Java methods and prevent boxing in loops. Tags are propagated bidirectionally
and the programmer is free to add them where they like.

core.typed uses tools.analyzer to obtain a rich abstract syntax tree that includes solved
reflection, however to tell the full story we include a rewrite relation 
in figure~\ref{figure:rewrite} to eliminate reflection when sufficient type hints are provided.
Figure~\ref{figure:hintinfer} models how type hints are propagated.

We implicitly resolve reflection before type checking and remaining reflective calls are type errors.

@@

\input{constant}
